---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Sophia Bryson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1 - WORKSPACE SETUP

# check working directory
getwd() 
  # When I run this before knitting, it's the base folder. 
  # Whenever I go to knit, it's the assignments folder. 
  # Using setwd() doesn't fix it. 

# load packages
library(tidyverse)
library(lubridate)
library(agricolae)
library(wesanderson) #y'know, for the fun colors. As always. 

# load data
LakesData <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", 
                      stringsAsFactors = TRUE) %>%
             mutate(sampledate = as.Date(sampledate, format = "%m/%d/%y"))


#2 - BUILT & SET GGPLOT THEME

  custom_theme <- theme_minimal(base_size = 11, base_family = "sans") +
                  theme(panel.background = element_rect(fill = "ivory2"),
                        panel.grid.major = element_line(color = "ivory3"),
                        panel.grid.minor = element_line(color = "ivory"),
                        axis.text = element_text(color = "ivory4"),
                        legend.position = "bottom",
                        plot.title = element_text(hjust = 0.5, family = "sans")) 

  #Set as default
  theme_set(custom_theme)


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across all lakes.
Ha: Mean lake temperature recorded during July does change with depth across all lakes. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4 - WRANGLE DATA 
JulyTempData <- LakesData %>% filter(lubridate::month(sampledate) == 7) %>%
                              select(lakename, year4, daynum, depth, temperature_C) %>%
                              na.omit()


#5
ggplot(JulyTempData, aes(x = temperature_C, y = depth)) + 
  geom_point(alpha = 0.2, color = wes_palettes$Cavalcanti1[4]) +
  geom_smooth(method = lm, se = FALSE, color = wes_palettes$Rushmore1[3]) +
  xlim(0, 35) +
  scale_y_reverse() +
  labs(x = "Temperature (C)", y = "Depth (m)", 
       title = "July Temperatures by Depth of NTL-LTER Lakes")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that there is a strong negative correlation between depth and temperature (though it looks positive when displayed on an inverted axis). As depth increases, temperature decreases. However, the sampe of the points suggests that this trend is non-linear and the relationship is not uniform across all depths. Rather, the change in temperature relative to depth is concentrated in the thermocline/metalimnion. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
(TempDepth_regression <- lm(data = JulyTempData, formula = temperature_C ~ depth))

summary(TempDepth_regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: This model indicates a moderate negative relationship between temperature and depth, predicting a decrease in temperature by 1.946 degrees Celcius for every 1m increase in depth. 73.9% of the variability in temperature is explained by changes in depth as indicated by the R-squared value, based on 9726 degrees of freedom. This result is significant to a p-value of less than 0.001. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 - CONSTRUCT AND CONDUCT AIC

tempAIC <- lm(data = JulyTempData, 
              formula = temperature_C ~ depth + year4 + daynum)

step(tempAIC)

#10 - CONGIFURE & RUN LM BASED ON AIC
#select lowest AIC (here, when none are dropped - use all)

(TempMulti_regression <- lm(formula = temperature_C ~ depth + year4 + daynum, 
                            data = JulyTempData))

summary(TempMulti_regression)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: the AIC method suggests retaining all of the provided variables (day, year, and depth) to predict temperature. This model explains 74.1% of the observed variance, only 0.2% more than the model using depth alone. Therefore, while technically an improvement, they gain is minimal, and the complexity of the model is increased to an extent perhaps unwarranted by the gain. 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
Temperature_ANOVA1 <- aov(data = JulyTempData, formula = temperature_C ~ lakename)
summary(Temperature_ANOVA1)

Temperature_ANOVA2 <- lm(data = JulyTempData, formula = temperature_C ~ lakename)
summary(Temperature_ANOVA2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference at the 0.001 level in the mean temperature between lakes. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(JulyTempData, aes(x = temperature_C, y = depth, color = lakename)) + 
  geom_point(alpha = 0.2) +
  geom_smooth(method = lm, se = FALSE, size = .4) +
  xlim(0, 35) +
  scale_y_reverse(limits = c(18, 0)) +
  scale_color_manual(values = c(wes_palettes$Cavalcanti1, wes_palettes$GrandBudapest1)) +
  labs(x = "Temperature (C)", y = "Depth (m)", 
       title = "July Temperatures by Depth of NTL-LTER Lakes")

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
# compute Tukey Honest Significant DIfferences
TukeyHSD(Temperature_ANOVA1)

# extract groupings
Temperature.groups <- HSD.test(Temperature_ANOVA1, "lakename", group = TRUE)
Temperature.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Statistically speaking, the mean July temperatures of Ward Lake and Paul Lake are the same as that of Peter Lake. No lake has a mean temperature that is statistically distinct from all other lakes: every lake is ggrouped with at leas one other lake. 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: To determine if the mean temperatures of a single pair of lakes (here, Peter and Paul) are significantly distinct from a statistical perspective, we could run a two-sample t-test, which allows for comparison of the means to two datasets. A two-sided t-test would tell us if the datasets are distinct. A one-sided t-test would allow us to test if one lakes was significantly higher or lower in mean July temperature than the other lake. 


